# Mediation with time-varying treatments, mediators, and covariates

```{r}
#| label: load-renv
#| echo: false
#| message: false
renv::load(here::here())
library(here)
```

We'll now turn to thinking about how to define and estimate effects
for cases where treatment, mediators, and covariates are time-varying.

Before we delve into the specific, here are a couple of papers on this
topic that are interesting/useful:

- [Mediation analysis with time varying exposures and
  mediators](https://rss.onlinelibrary.wiley.com/doi/full/10.1111/rssb.12194)
  by Tyler J. VanderWeele and Eric J. Tchetgen Tchetgen
- [Longitudinal Mediation Analysis with Time-varying Mediators and
  Exposures, with Application to Survival
  Outcomes](https://www.degruyter.com/document/doi/10.1515/jci-2016-0006/html)
  by  Wenjing Zheng and Mark J. van der Laan
- [Efficient and flexible causal mediation with time-varying
  mediators, treatments, and
  confounders](https://www.degruyter.com/document/doi/10.1515/jci-2022-0077/html) by Iván Díaz,
  Nicholas Williams, and Kara E. Rudolph
- [Identification and estimation of mediational effects of
  longitudinal modified treatment
  policies](https://arxiv.org/abs/2403.09928) by Brian Gilbert,
  Katherine L. Hoffman, Nicholas Williams, Kara E. Rudolph, Edward
  J. Schenck, Iván Díaz.

In this chapter, we will be using the [`lcm` `R`
package](https://github.com/nt-williams/lcm), which supports binary
time-varying treatments and categorical mediators, and the [`lcmmtp`
`R` package](https://github.com/nt-williams/lcmmtp) package, which
supports continuous, multivariate, and binary time-varying treatments
and categorical mediators.

## Illustrative example

As an illustrative example, we will use a dataset from an
observational study looking at the effect of invasive mechanical
ventilation (IMV) on the survival of COVID-19 patients, considering
acute kidney injury (AKI) as a mediating factor.

Briefly, IMV is a treatment for acute respiratory distress syndrome
(ARDS). While IMV is a potentially life-saving therapy in patients
with ARDS, its usage has been associated with several iatrogenic
risks. Of interest to our illustrative study is acute kidney injury (AKI),
a critical condition that complicates ICU stays and is associated with
increased mortality. The causal model underlying this problem is as
follows:

The data on a single observational unit can be represented by the
vector $O=(L_1,
A_1, Z_1, M_1, L_2, \ldots, A_\tau, Z_\tau, M_\tau, Y)$, with the data
pooled across all participants denoted $O_1, \ldots, O_n$, for a of
$n$ i.i.d. observations of $O$. The associated DAG is

```{tikz}
#| fig-cap: Directed acyclic graph for time-varying treatments, mediators, and confounders
\dimendef\prevdepth=0
\pgfdeclarelayer{background}
\pgfsetlayers{background,main}
\usetikzlibrary{arrows,positioning}
\tikzset{
>=stealth',
punkt/.style={
rectangle,
rounded corners,
draw=black, very thick,
text width=6.5em,
minimum height=2em,
text centered},
pil/.style={
->,
thick,
shorten <=2pt,
shorten >=2pt,}
}
\newcommand{\Vertex}[3]
{\node[minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {#3};
}
\pgfarrowsdeclare{arcs}{arcs}{...}
{
  \pgfsetdash{}{0pt} % do not dash
  \pgfsetroundjoin   % fix join
  \pgfsetroundcap    % fix cap
  \pgfpathmoveto{\pgfpoint{-5pt}{5pt}}
  \pgfpatharc{180}{270}{5pt}
  \pgfpatharc{90}{180}{5pt}
  \pgfusepathqstroke
}
\newcommand{\VertexR}[2]
{\node[rectangle, draw, minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {$#2$};
}
\newcommand{\ArrowR}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend right=30] (#2);
\end{pgfonlayer}
}
\newcommand{\ArrowB}[3]%
{ \begin{pgfonlayer}{background}
    \draw[|-arcs,line width=0.4mm,shorten <= 0.3cm,shorten >= 0.3cm,#3] (#1) -- +(#2);
  \end{pgfonlayer}
}
\newcommand{\ArrowL}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend left=45] (#2);
\end{pgfonlayer}
}
\newcommand{\EdgeL}[3]
{ \begin{pgfonlayer}{background}
\draw[dashed,#3] (#1) to[bend right=-45] (#2);
\end{pgfonlayer}
}
\newcommand{\Arrow}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) -- +(#2);
\end{pgfonlayer}
}
\begin{tikzpicture}
    \Vertex{-1, 0}{W}{$L_1$}
    \Vertex{1, 0}{A1}{$A_1$}
    \Vertex{2, 0}{M1}{$M_1$}
    \Vertex{3, 0}{Z1}{$L_2$}
    \Vertex{2, -1}{L1}{$Z_1$}

    \ArrowB{W}{A1}{black}

    \Arrow{A1}{M1}{black}
    \Arrow{A1}{L1}{black}
    \Arrow{M1}{Z1}{black}
    \Arrow{L1}{M1}{black}
    \Arrow{L1}{Z1}{black}
    \ArrowL{A1}{Z1}{black}

    \Vertex{5, 0}{A2}{$A_2$}
    \Vertex{6, 0}{M2}{$M_2$}
    \Vertex{7, 0}{Z2}{$L_3$}
    \Vertex{6, -1}{L2}{$Z_2$}
    \node (dots) at (8, 0) {$\cdots$};
    \Arrow{A2}{M2}{black}
    \Arrow{A2}{L2}{black}
    \Arrow{M2}{Z2}{black}
    \Arrow{L2}{M2}{black}
    \Arrow{L2}{Z2}{black}
    \ArrowL{A2}{Z2}{black}

    \ArrowB{Z1}{A2}{black}

    \Vertex{9, 0}{At}{$A_\tau$}
    \Vertex{10, 0}{Mt}{$M_\tau$}
    \Vertex{11, 0}{Zt}{$Y$}
    \Vertex{10, -1}{Lt}{$Z_\tau$}
    \Arrow{At}{Mt}{black}
    \Arrow{At}{Lt}{black}
    \Arrow{Mt}{Zt}{black}
    \Arrow{Lt}{Mt}{black}
    \Arrow{Lt}{Zt}{black}
    \ArrowL{At}{Zt}{black}
\end{tikzpicture}
```

Where we are using the following notation:

- $W$: baseline variables such as comorbidities, demographics, etc.
- $L_t$ and $Z_t$: time-varying covariates such as lab results,
  vitals, treatments, etc.
- $A_t$: type of oxygen support at time $t$ (0: no oxygen support, 1:
  oxygen support excluding IMV, 2: IMV)
- $M_t$: indicator of AKI at time $t$
- $Y$: mortality at end of study
- We will use $H_t$ as a shorthand to denote all the data measured up
  until right before $A_t$ occurs

## Defining causal effects in this example

How can we define (total) causal effects in this example?

- Main challenge: cannot consider static treatmemt regimes (e.g., do
  not intubate)
- Such regimes would not be supported in the data (doctors would always intubate a person whose blood oxygen is too low)
- We use *modified treatment policies* to address this
- Main idea: consider a slight modification to the treatment a patient
  actually received
- For example, can consider the effect of a small delay in receiving
  IMV
- Specifically, we will consider a delay of one day in receiving IMV
- In notation, the treatment regime would be as follows:

\begin{equation}
  d_t(a_t,h_t) =
  \begin{cases}
    1 &\text{ if } a_t=2 \text{ and } a_s \leq 1 \text{ for all } s < t,\\
    a_t & \text{ otherwise.}
  \end{cases}
\end{equation}
- We could then define the total effect as $E[Y(d) - Y]$, where $Y(d)$
  is the counterfactual mortality if the above rule had been
  implemented every day, i.e., the intervention is $d=(d_1,d_2,\ldots,
  d_\tau)$.
- This is a contrast of the mortality rate under a treatment rule that would
  delay intubation by one day vs the mortality rate that was actually observed.

## How do we define mediation causal effects with time-varying data?

The above causal effect could be decomposed into natural direct and
indirect effects as follows

\begin{align*}
E[Y(d) - Y] & = E[Y(d, M(d)) - Y(A, M)]\\
&=\underbrace{\E[Y(\color{red}{d},\color{blue}{M(d)}) -
    Y(\color{red}{d},\color{blue}{M})]}_{\text{natural indirect effect}} +
    \underbrace{\E[Y(\color{blue}{d},\color{red}{M}) -
    Y(\color{blue}{A},\color{red}{M})]}_{\text{natural direct effect}}
\end{align*}

- However, as before, these natural mediation effects are not identified.

- The reason is that time-varying mediators exacerbate the issue of intermediate
confounding. To see why, let us look at the DAG again:

```{tikz}
#| fig-cap: Directed acyclic graph with intermediate confounding
\dimendef\prevdepth=0
\pgfdeclarelayer{background}
\pgfsetlayers{background,main}
\usetikzlibrary{arrows,positioning}
\tikzset{
>=stealth',
punkt/.style={
rectangle,
rounded corners,
draw=black, very thick,
text width=6.5em,
minimum height=2em,
text centered},
pil/.style={
->,
thick,
shorten <=2pt,
shorten >=2pt,}
}
\newcommand{\Vertex}[3]
{\node[minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {#3};
}
\pgfarrowsdeclare{arcs}{arcs}{...}
{
  \pgfsetdash{}{0pt} % do not dash
  \pgfsetroundjoin   % fix join
  \pgfsetroundcap    % fix cap
  \pgfpathmoveto{\pgfpoint{-5pt}{5pt}}
  \pgfpatharc{180}{270}{5pt}
  \pgfpatharc{90}{180}{5pt}
  \pgfusepathqstroke
}
\newcommand{\VertexR}[2]
{\node[rectangle, draw, minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {$#2$};
}
\newcommand{\ArrowR}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend right=30] (#2);
\end{pgfonlayer}
}
\newcommand{\ArrowB}[3]%
{ \begin{pgfonlayer}{background}
    \draw[|-arcs,line width=0.4mm,shorten <= 0.3cm,shorten >= 0.3cm,#3] (#1) -- +(#2);
  \end{pgfonlayer}
}
\newcommand{\ArrowL}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend left=45] (#2);
\end{pgfonlayer}
}
\newcommand{\EdgeL}[3]
{ \begin{pgfonlayer}{background}
\draw[dashed,#3] (#1) to[bend right=-45] (#2);
\end{pgfonlayer}
}
\newcommand{\Arrow}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) -- +(#2);
\end{pgfonlayer}
}
\begin{tikzpicture}
    \Vertex{-1, 0}{W}{$L_1$}
    \Vertex{1, 0}{A1}{$A_1$}
    \Vertex{2, 0}{M1}{$M_1$}
    \Vertex{3, 0}{Z1}{$L_2$}
    \Vertex{2, -1}{L1}{$Z_1$}

    \ArrowB{W}{A1}{black}

    \Arrow{A1}{M1}{black}
    \Arrow{A1}{L1}{black}
    \Arrow{M1}{Z1}{black}
    \Arrow{L1}{M1}{black}
    \Arrow{L1}{Z1}{black}
    \ArrowL{A1}{Z1}{black}

    \Vertex{5, 0}{A2}{\textcolor{teal}{$A_2$}}
    \Vertex{6, 0}{M2}{\textcolor{orange}{$M_2$}}
    \Vertex{7, 0}{Z2}{\textcolor{orange}{$L_3$}}
    \Vertex{6, -1}{L2}{\textcolor{orange}{$Z_2$}}
    \node (dots) at (8, 0) {$\cdots$};
    \Arrow{A2}{M2}{black}
    \Arrow{A2}{L2}{black}
    \Arrow{M2}{Z2}{black}
    \Arrow{L2}{M2}{black}
    \Arrow{L2}{Z2}{black}
    \ArrowL{A2}{Z2}{black}

    \ArrowB{Z1}{A2}{black}

    \Vertex{9, 0}{At}{\textcolor{orange}{$A_\tau$}}
    \Vertex{10, 0}{Mt}{\textcolor{violet}{$M_\tau$}}
    \Vertex{11, 0}{Zt}{\textcolor{violet}{$Y$}}
    \Vertex{10, -1}{Lt}{\textcolor{orange}{$Z_\tau$}}
    \Arrow{At}{Mt}{black}
    \Arrow{At}{Lt}{black}
    \Arrow{Mt}{Zt}{black}
    \Arrow{Lt}{Mt}{black}
    \Arrow{Lt}{Zt}{black}
    \ArrowL{At}{Zt}{black}
\end{tikzpicture}
```
Note that all the variables in orange are confounders of the
  mediator $M_\tau$ and the outcome, and are
  also affected by treatment at time $t=2$.

One possible solution to the above issues involves considering
randomized versions of the above effects. Specifically:

- Define $G(d)$ to be a random draw from the distribution of $M(d)$
  conditional on baseline variables $W$.
- We can then obtain the decomposition
  $$
    \E[Y(d, G(d)) - Y(A, G(A))]=\underbrace{\E[Y(\color{red}{d},\color{blue}{G(d)}) -
    Y(\color{red}{d},\color{blue}{G(A)})]}_{\text{randomized interventional indirect effect}} +
    \underbrace{\E[Y(\color{blue}{d},\color{red}{G(A)}) -
      Y(\color{blue}{A},\color{red}{G(A)})]}_{\text{randomized interventional
        direct effect}}
  $$
- As an example, consider the counterfactual $Y(d, G(d))$.
- $M(d)$ is the observed AKI status of patients under a delay in intubation. If
  $W$ is age, and we are deciding how to intervene on a patient who is 45 years
  old, we take all the AKI statuses of 45 year olds and draw one of these AKI
  values at random. Call this random draw $G(d)$
- For a 45 year old patient, $Y(d, G(d))$ is the counterfactual mortality of
  a patient if intubation had been delayed, and their AKI status would have
  been assigned to a random draw from the AKI status of 45 year patients.

## Identification assumptions and formula

The above effects are identified under the following assumptions:

- All the common causes of $A_t$ and $(Z_s, M_s, A_{s+1}, L_{s+1})$ are
  measured for $s\geq t$
- All the common causes of $M_t$ and $(Z_{s+1}, A_{s+1}, L_{s+1})$ are measured
  for $s\geq t$
- The intervention $d$ is supported in the data, meaning that for every patient
  with covariates $h_t$ who had treatment status $a_t$, it is possible to find
  a patient with covariates $h_t$ who had treatment status $d(a_t, h_t)$
  - In our example, this translates roughly as: for every patient with
    covariate history $h_t$ who was intubated at time $t$, it is possible to
    find a patient covariate history $h_t$ who was intubated at time $t+1$.
- There is a positive probability of the mediator $M_t$ for all feasible
  covariate histories.

The identification formula is complex, but we will explain it in the case of
two time points. That is, assume the data are $O=(W, A_1, Z_1, M_1, L_1, A_2,
Z_2, M_2, Y)$.

Identification can be based on the following procedure. First, for each value
$m_1$ and $m_2$ of the mediator, compute outcome regressions as follows:

1. Regress $Y$ on $W, A_1, Z_1, M_1, L_1, A_2, Z_2, M_2$. Use this
   regression to predict the outcome had the mediator been set to
   $M_2=m_2$. Let $\tilde Y_2(m_2)$ denote this prediction.
2. Regress $\tilde Y_2(m_2)$ on $W, A_1, Z_1, M_1, L_1, A_2$. Use this
   regression to predict the outcome had the treatment $A_2$ been set to
   $d(A_2, H_2)$. Let $\tilde Y_2^d(m_2)$ denote this prediction.
3. Regress $\tilde Y_2(m_2, d_2)$ on $W, A_1, Z_1, M_1$. Use this
   regression to predict the outcome had the mediator been set to
   $M_1=m_1$. Let $\tilde Y_1(m_1, m_2)$ denote this prediction.
4. Regress $\tilde Y_1(m_1, m_2)$ on $W, A_1$. Use this
   regression to predict the outcome had the treatment $A_1$ been set to
   $d(A_1, H_1)$. Let $\tilde Y_1^d(m_1, m_2)$ denote this prediction.

Then, for each value $m_1$ and $m_2$ of the mediator, compute the mediator
distribution as follows:

1. Regress the binary variable $I(M_2=m_2)$ on $W, A_1, Z_1, M_1, L_1, A_2$.
   Use this model to predict the probability of $M_2=m_2$ under an intervention
   that sets $A_2$ to $d(A_2, H_2)$. Let this predicted probability be denoted
   with $P(m_2)$.
2. Regress the binary variable $I(M_1=m_1)P(m_2)$ on $W, A_1$. Use this model
   to predict under an intervention that sets $A_1$ to $d(A_1, H_1)$. Let this
   prediction be denoted with $P(m_1, m_2)$.

At the end of these two sequential regression procedures, we have
values $\tilde Y_1^d(m_1, m_2)$ and $P(m_1, m_2)$ for each value of
the mediator $(m_1, m_2)$. Then, under identification assumptions, we
have:

$$
  \E[Y(d, G(d)) = \sum_{m_1, m_2}\tilde Y_1^d(m_1, m_2)P(m_1, m_2) \ .
$$

## Estimators and `R` package

As before, we can develop inverse probability weighted estimators, as well as
substitution estimators based on the g-computation formula and doubly robust
(DR) estimators.

All of these estimators get significantly more complex. For instance, an
g-computation estimator may be developed by running the regressions indicated
in the above sequential regression procedures.

Fortunately, the doubly robust estimators are coded in a package that
can be used off-the-shelf without having to code any complicated
sequential regression strategies on your own. Let us look at an
example from the `lcmmtp` R package. First, let's take a look at a
simulated dataset available in the package:

```{r}
#| label: longit-data
library(lcmmtp)
data <- as.data.frame(apply(lcmmtp_foo, 2, as.numeric))
head(data)
dim(lcmmtp_foo)
```

Now, let us perform an analysis where we assume our intent is to estimate
$\E[Y(1), G(0)]$:

```{r}
#| label: longit-analysis1
library(mlr3extralearners)
vars <- lcmmtp:::lcmmtp_variables$new(
    L = list(c("L_1"), c("L_2")),
    A = c("A_1", "A_2"),
    Z = list(c("Z_1"), c("Z_2")),
    M = c("M_1", "M_2"),
    Y = "Y",
    cens = c("c1", "c2")
)
lrnrs <- c("mean", "earth", "glm")
d_ap <- function(data, trt) rep(1, length(data[[trt]]))
d_as <- function(data, trt) rep(0, length(data[[trt]]))

EY10 <- lcmmtp(
  data, vars, d_ap, d_as,
  control = .lcmmtp_control(folds = 2,
                            learners_trt = lrnrs,
                            learners_mediator = lrnrs,
                            learners_QL = lrnrs,
                            learners_QZ = lrnrs,
                            learners_QM = lrnrs)
)
```

Now, assume that we want to estimate the direct effect by contrasting
$\E[Y(1), G(0)] - \E[Y(0), G(0)]$:

```{r}
#| label: longit-analysis2
EY00 <- lcmmtp(
  lcmmtp_foo, vars, d_as, d_as,
  control = .lcmmtp_control(folds = 2,
                            learners_trt = lrnrs,
                            learners_mediator = lrnrs,
                            learners_QL = lrnrs,
                            learners_QZ = lrnrs,
                            learners_QM = lrnrs)
)
```

And we can contrast the two using a convenient function from the `lmtp`
R package:

```{r}
#| label: contrast
library(lmtp)
class(EY00) <- class(EY10) <- 'lmtp'
EY00$estimator <- EY10$estimator <- 'SDR'
names(EY00)[3] <- names(EY10)[3] <- 'eif'
lmtp_contrast(EY10, ref = EY00)
```

## Pros and cons of this methodology

#### Pros {.unnumbered}

- Allows the non-parametric definition, identification, and estimation of
  mediational causal effects for general longitudinal data structures
- Allows for the use of machine learning to alleviate model
  misspecification bias, and is equipped with formulas for the
  computation of correct standard errors and confidence intervals
- Easy-to-use software

#### Cons {.unnumbered}

- Some limitations remain: mediators $M$ need to be discrete random variables
- As before, interventional effects do not satisfy the mediational sharp null
  criteria, meaning that they may be different from zero when no individual in
  the population experiences mediational effects
  - This is probably not a big worry in practice, but it is something
    we are keeping in mind as we develop novel estimators
